### 2015-04-16 manuscript/Text_S1.tex
Commit prior to PLoS One submission

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{A\arabic{figure}}
\renewcommand\thetable{A\arabic{table}}
\renewcommand\theequation{A.\arabic{equation}}

@@

\begin{center} \bf {\large On extrapolating past the range of observed data when making statistical predictions in ecology}

\vspace{0.7cm}
Paul B. Conn$^{1*}$, Devin S. Johnson$^1$, and Peter L. Boveng$^1$
\end{center}
\vspace{0.5cm}

\rm
\small

\it $^1$National Marine Mammal Laboratory, Alaska Fisheries Science Center,
NOAA National Marine Fisheries Service,
Seattle, Washington 98115 U.S.A.\\

@@

\begin{center}
Appendix A: Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model
\bigskip
\end{center}
\vspace{.3in}

@@

\hspace{.5in}The RSR approach to spatial regression uses a reduced-rank version of the popular intrinsic conditionally autoregressive \citep[ICAR;][]{Besag1995,RueHeld2005} model for spatial random effects, reparameterized so that basis vectors are orthogonal to the main effects of interest.  This approach has generated substantial recent interest, as fixed effects retain primacy in explaining variation in the ecological process of interest and problems with spatial confounding between fixed and random effects are eliminated. As such, spatial random effects are only used to account for residual autocorrelation \citep{Reich2006,Hodges2010} and the decision to incorporate spatial autocorrelation has little effect on the point estimates of fixed effects.  In addition, reduced dimension spatial models such as RSR lighten computational burden while still accounting for course-scale spatial autocorrelation \citep[see e.g.][]{LatimerEtAl2009,Wikle2010,Hughes2013}.  It turns out that RSR models can also be written in the form of Eq. \ref{eq:PC}, using a
different choice of ${\bf K}$ \citep{Hughes2013}, constructed as follows:
\begin{enumerate}
  \item Define the residual projection matrix ${\bf P}^\perp={\bf I}-{\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$
  \item Calculate the Moran operator matrix $\boldsymbol{\Omega}=J{\bf P}^\perp {\bf W}{\bf P}^\perp/{\bf 1}^\prime {\bf W} {\bf 1}$, where $J$ is the number of areal survey units
  \item Define ${\bf K}$ as an $(n \times m)$ matrix, where the columns of ${\bf K}$ are composed of the eigenvectors  associated with the largest $m$ eigenvalues of $\boldsymbol{\Omega}$. Hughes and Haran (\citeyear{Hughes2013}) used simulation to explore a range of such values and concluded $m=50-100$ should suffice for most applications.
\end{enumerate}
Here, ${\bf I}$ an identity matrix, ${\bf 1}$ is a column vector of ones, and ${\bf W}$ represents an association matrix describing the spatial neighborhood structure of sampling units.  For instance, for a first order neighborhood structure, ${\bf W}$ would include a 1 for all rows $i$ and columns $j$ where sampling unit $i$ and $j$ are neighbors \citep[see][for alternative association matrices]{RueHeld2005}.

@@

Although $\boldsymbol{\mu}$ could potentially be integrated out of the likelihood, its inclusion allows for easier computation of full conditionals for $\boldsymbol{\beta}$ and $\tau_\epsilon$ (and for other full conditionals once the model is expanded, e.g. for GAMs and STRMs).  During the main MCMC phase, the $\boldsymbol{\mu}$ only need to be updated for grid cells that are exposed to sampling (see a subsequent section for information on posterior prediction for details on unsampled areas).  The full conditional for $\mu_s$ in a given grid cell $s$ is given by
\begin{equation*}
  [\mu_{s,t} | \cdot] \propto \mathcal{N}(\mu_s;{\bf X}_s \boldsymbol{\beta},\tau_\epsilon^{-0.5}) \times \textrm{Poisson}(C_s,\exp(o_s+\mu_s)),
\end{equation*}
where ${\bf X}_s$ gives the $s$th row of the design matrix and $o_s$ is an offset.  For our examples, quadrats are configured to represent $10\%$ of a grid cell, cell $o_s = \log(0.1)$.
We use a Metropolis-Hastings step to sample from $[\mu_{s,t} | \cdot]$.

@@

In order to calculate prediction variance on the real scale for generalized models, it is necessary to use a procedure such as the delta method \citep{Dorfman1938}.  In particular, we can write the variance of the Poisson expectation as
\begin{linenomath*}
\begin{eqnarray}
  \label{eq:var-lambda}
  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
%      \vdots & \vdots & \ddots & \vdots \\
%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
%  \end{array} \right].
\end{eqnarray}
\end{linenomath*}
where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).  Under common univariate link functions (e.g. log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g. multinomial logit) $\boldsymbol{\Delta}$ will be dense.

@@

\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and substitute the posterior distribution $[\boldsymbol{\theta}|{\bf Y}]$ for $[\boldsymbol{\theta}]$ in Eq. \ref{eq:E_var_lambda}. Under this approach, one simply sets ${\rm Var}(\mu_x)$ equal to the posterior variance of $\mu_x$ (across Markov chain iterations).  In the case of the log link function, one also needs an estimate for $\mu_x$ to use in delta method calculations; we used the posterior median for each sampling unit.  Although we use a Bayesian formulation for modeling convenience, one could also consider substituting parametric approximations for ${\rm Var}(\mu_x)$ and $\mu_x$ based on output from standard likelihood-based analyses.

>>>>

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{S1\arabic{figure}}
\renewcommand\thetable{S1\arabic{table}}
\renewcommand\theequation{S1.\arabic{equation}}

@@

@@

\begin{center}
S1 Text: Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model
\bigskip
\end{center}
\vspace{.3in}

@@

\hspace{.5in}The RSR approach to spatial regression uses a reduced-rank version of the popular intrinsic conditionally autoregressive \citep[ICAR;][]{Besag1995,RueHeld2005} model for spatial random effects, reparameterized so that basis vectors are orthogonal to the main effects of interest.  This approach has generated substantial recent interest, as fixed effects retain primacy in explaining variation in the ecological process of interest and problems with spatial confounding between fixed and random effects are eliminated. As such, spatial random effects are only used to account for residual autocorrelation \citep{Reich2006,Hodges2010} and the decision to incorporate spatial autocorrelation has little effect on the point estimates of fixed effects.  In addition, reduced dimension spatial models such as RSR lighten computational burden while still accounting for course-scale spatial autocorrelation \citep[see e.g.][]{LatimerEtAl2009,Wikle2010,Hughes2013}.  This approach works by reformulating the spatial random effects in Eq. \ref{eq:spat_reg} as $\boldsymbol{\eta} = {\bf K} \boldsymbol{\alpha}$, where $\alpha$ is a vector of $m$ random effects in a reduced dimension subspace.
The matrix ${\bf K}$ can be constructed as follows \citep{Hughes2013}:
\begin{enumerate}
  \item Define the residual projection matrix ${\bf P}^\perp={\bf I}-{\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$
  \item Calculate the Moran operator matrix $\boldsymbol{\Omega}=J{\bf P}^\perp {\bf W}{\bf P}^\perp/{\bf 1}^\prime {\bf W} {\bf 1}$, where $J$ is the number of areal survey units
  \item Define ${\bf K}$ as an $(n \times m)$ matrix, where the columns of ${\bf K}$ are composed of the eigenvectors  associated with the largest $m$ eigenvalues of $\boldsymbol{\Omega}$. Hughes and Haran (\citeyear{Hughes2013}) used simulation to explore a range of such values and concluded $m=50-100$ should suffice for most applications.
\end{enumerate}
Here, ${\bf I}$ an identity matrix, ${\bf 1}$ is a column vector of ones, and ${\bf W}$ represents an association matrix describing the spatial neighborhood structure of sampling units.  For instance, for a first order neighborhood structure, ${\bf W}$ would include a 1 for all rows $i$ and columns $j$ where sampling unit $i$ and $j$ are neighbors \citep[see][for alternative association matrices]{RueHeld2005}.

@@

Although $\boldsymbol{\mu}$ could potentially be integrated out of the likelihood, its inclusion allows for easier computation of full conditionals for $\boldsymbol{\beta}$ and $\tau_\epsilon$ (and for other full conditionals once the model is expanded, for example, to GAMs and STRMs).  During the main MCMC phase, the $\boldsymbol{\mu}$ only need to be updated for grid cells that are exposed to sampling (see a subsequent section for information on posterior prediction for details on unsampled areas).  The full conditional for $\mu_s$ in a given grid cell $s$ is given by
\begin{equation*}
  [\mu_{s,t} | \cdot] \propto \mathcal{N}(\mu_s;{\bf X}_s \boldsymbol{\beta},\tau_\epsilon^{-0.5}) \times \textrm{Poisson}(C_s,\exp(o_s+\mu_s)),
\end{equation*}
where ${\bf X}_s$ gives the $s$th row of the design matrix and $o_s$ is an offset.  For our examples, quadrats are configured to represent $10\%$ of a grid cell, cell $o_s = \log(0.1)$.
We use a Metropolis-Hastings step to sample from $[\mu_{s,t} | \cdot]$.

@@

In order to calculate prediction variance on the real scale for generalized models, one can use a procedure such as the delta method \citep{Dorfman1938}, or in the case of Bayesian inference, can simply calculate posterior predictive variance of $\boldsymbol{\lambda}_x$.  For the delta method approach, we can write the variance of the Poisson expectation as
\begin{linenomath*}
\begin{eqnarray}
  \label{eq:var-lambda}
  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
%      \vdots & \vdots & \ddots & \vdots \\
%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
%  \end{array} \right].
\end{eqnarray}
\end{linenomath*}
where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).  Under common univariate link functions (e.g. log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g. multinomial logit) $\boldsymbol{\Delta}$ will be dense.

@@

\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and to use posterior predictive inference to summarize ${\rm var}(\boldsymbol{\lambda}_x | {\bf Y})$.  Although we use a Bayesian formulation for modeling convenience, one could also consider substituting parametric approximations for ${\rm Var}(\mu_x)$ and $\mu_x$ based on output from standard likelihood-based analyses.

### 2015-04-16 manuscript/Text_S2.tex
Commit prior to PLoS One submission

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{B\arabic{figure}}
\renewcommand\thetable{B\arabic{table}}
\renewcommand\theequation{B.\arabic{equation}}

@@

\begin{center}
Appendix B: Full details of simulation study examining predictive extrapolation
\bigskip
\end{center}
\vspace{.3in}

@@

Spatially explicit statistical models are increasingly used to estimate animal abundance and predict species distributions from count data.  We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where predictions of animal abundance using these models may be problematic, and to determine whether statistical inferences are more robust when restricted to sampling units within the gIVH.  Each simulation replicate consisted of several steps (Fig. \ref{}), including
\begin{enumerate}
  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30 by 30 grid,
  \item Simulate animal abundance across the landscape as a function of environmental covariates,
  \item Simulate animal population surveys across the landscape, including the position of count quadrats and the resulting animal counts,
  \item Estimate animal abundance as a function of two of the environmental covariates according three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).  For the latter, only a spatial dimension was modeled.  All such models were specified hierarchically, with MCMC used for posterior simulation.
  \item Calculate the gIVH using realized posterior variance (i.e. after data were collected and analyzed).
\end{enumerate}
We now describe each of these tasks and results of the simulation study in further detail before describing results.  All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the R package \texttt{SpatPred} that accompanies this article.

@@

For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models, corresponding to a generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a generalized additive model (GAM; smooth effects of covariates on the log scale), and a spatio-temporal regression model (with both linear, fixed effects and spatially autocorrelated random effects).  Further details on these models, together with the procedure used for posterior simulation, are presented in Appendix A.  For this study, we used the RSR implementation of the STRM outlined in Appendix A.  Prior distributions for each precision parameter, $\tau$, were set to $\text{Gamma}(1.0,0.01)$, which is diffuse while maintaining a flat shape near the origin.  Regression parameters ($\beta$) were given vague $\mathcal{N}(0,\tau=0.01)$ priors.

@@

As suggested in Appendix A, we used Eq. A.7 to calculate the gIVH, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.

\centerline{\sc Results}

Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig. \ref{fig:sim_boxplots}).  This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.  The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.  Interestingly, the convolution kernel GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.  For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude (e.g. $\approx 0.3$; Fig. \ref{fig:sim_boxplots}) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{sim_maps.pdf}
\end{center}
\caption{Depiction of a single simulation scenario.  Panels (A-C) give simulated covariate values, panel D gives true animal abundance, (E) gives estimated abundance from a GLM run on count data from a spatially balanced survey design, and (F) gives abundance from a GLM applied to count data from a convenience survey.  In (E-F), predictions outside the gIVH are represented by black boxes, and sampling locations are represented with an x.  For the convenience sample, the median posterior abundance prediction for the entire survey area is $57\%$ greater than true abundance when inference is made to the whole study area.  When inference is limited to the gIVH, median posterior abundance was just $16\%$ greater than true abundance.}
\label{fig:sim_maps}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{bias_generic.pdf}
\end{center}
\caption{Boxplots summarizing distribution of proportional error in the posterior predictive median of abundance for the simulation study as a function of estimation model (x-axis), survey design (columns) and whether or not inference was restricted to the gIVH (rows).  The lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.  Outliers outside of this range are denoted with points.  Horizontal lines within boxes denote median bias.  The two numbers located below each boxplot indicate mean bias (upper number) and the number of additional outliers for which proportional bias was greater than 2.0 (lower number). }
\label{fig:sim_boxplots}
\end{figure}

>>>>

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{S2.\arabic{figure}}
\renewcommand\thetable{S2.\arabic{table}}
\renewcommand\theequation{S2.\arabic{equation}}

@@

\begin{center}
S2 Text: Full details of simulation study examining predictive extrapolation
\bigskip
\end{center}
\vspace{.3in}

@@

Spatially explicit statistical models are increasingly used to estimate animal abundance and predict species distributions from count data.  We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where predictions of animal abundance using these models may be problematic, and to determine whether statistical inferences are more robust when restricted to sampling units within the gIVH.  Each simulation replicate consisted of several steps, including
\begin{enumerate}
  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30$\times$30 grid,
  \item Simulate animal abundance across the landscape as a function of environmental covariates,
  \item Simulate animal population surveys across the landscape, including the position of count quadrats and the resulting animal counts,
  \item Estimate animal abundance as a function of two of the environmental covariates according three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).  For the latter, only a spatial dimension was modeled.  All such models were specified hierarchically, with MCMC used for posterior simulation.
  \item Calculate the gIVH using realized posterior variance (i.e. after data were collected and analyzed).
\end{enumerate}
We now describe each of these tasks and results of the simulation study in further detail before describing results.  All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the R package \texttt{SpatPred} that accompanies this article.

@@

For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models, corresponding to a generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a generalized additive model (GAM; smooth effects of covariates on the log scale), and a spatio-temporal regression model (with both linear, fixed effects and spatially autocorrelated random effects).  Further details on these models, together with the procedure used for posterior simulation, are presented in Text S1.  For this study, we used the RSR implementation of the STRM outlined in Text S1.  Prior distributions for each precision parameter, $\tau$, were set to $\text{Gamma}(1.0,0.01)$, which is diffuse while maintaining a flat shape near the origin.  Regression parameters ($\beta$) were given vague $\mathcal{N}(0,\tau=0.01)$ priors.

@@

As suggested in Text S1, we used Eq. S1.7 to calculate the gIVH, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.

@@

@@
